Astrophysical calibration of gravitational-wave detectors

### 2015-08-06 docs/cosmic_cal.tex
Added more to the document - still need to go through some more bits and check equation 6 is ok

\begin{abstract}
  We present an investigation into the potential of calibrating \MP{change to: ``assessing the validity of the calibration''?}
  gravitational wave detector outputs through the use of standard
  sirens. Such signals, as measured via gravitational wave
  observations provide an estimated luminosity distance which is
  subject to uncertainties in the calibration of the data.  If a host
  galaxy is identified for a given source then its redshift can be
  obtained and combined with current knowledge of the cosmological
  parameters.  This will yield the true luminosity distance and allow
  the direct comparison with the estimated value.  Discrepancies can
  then be used to correct the original calibration.~\cm{Update at end}
\end{abstract}

@@

% introduce GWs and BNS systems
It is expected that the advanced generation of interferometric \ac{GW}
detectors will detect waves emitted from $O(10s)$ of compact binary
coalescences.  One such class of these cataclysmic events, the inspiral and
merger of \ac{BNS} systems will be detected out to a maximum range of $\approx
450$ Mpc. Assuming the current best estimates for the cosmological parameters,
this is equivalent to a redshift $z\approx 0.1$. As noted by \cite{1986Natur.323..310S} \ac{CBC}
systems can be used as cosmological distance markers, otherwise known as
``standard sirens'' (analogous to the \ac{EM} standard candles). Standard sirens
allow ...

@@

% talk about the calibration idea
If a single \ac{GW} event is observed in coincidence with a \ac{sGRB} it will \MP{may?}
be possible to identify the host galaxy of the \ac{BNS} source.  With this
information a spectroscopic redshift can be very accurately obtained. With
knowledge of the redshift, using current best estimates for the Hubble constant
and other cosmological parameters, the true luminosity distance can be
estimated to $\sim 1\%$ accuracy.  The \ac{GW} measurement acts as a standard
siren also giving us a direct measurement of the luminosity distance to the source.
The accuracy of such a measurement depends on a number of factors including the
accuracy with which the \ac{GW} detector has been calibrated.  Hence comparison
with the distance estimate from the \ac{sGRB} we can re-calibrate (or validate)
the existing experimentally obtained calibration.

% What do we do in this paper
In this paper we investigate the feasibility of this approach for single
coincident \ac{GW}--\ac{sGRB} events and establish the validation power of such
a calibration technique.  In Sec.~\ref{sec:calibration} we briefly summarize
the existing experimental technique for \ac{GW} detector calibration and its
expected accuracy.  We then review the concept of \ac{GW} standard sirens and
their proposed coincident \ac{EM} signatures, the \acp{sGRB} in
Secs.~\ref{sec:sirens} and~\ref{sec:GRB}. In Sec.~\ref{sec:single} we present
the results of Monte-Carlo simulations for the purpose of luminosity distance
estimation using \ac{GW} events. In Sec.~\ref{sec:cosmo} we show the likely
accuracy of \ac{sGRB} distances obtained from their host galaxy redshift and in
Sec.~\ref{sec:discussion} we conclude.

@@

\cm{We need to define our model for the calibration, i.e. are we talking
about amplitude only and are we breaking this up over frequency.}

\MP{In this paper I think we stick to just talking about a single overall amplitude calibration scale.}

The technique used throughout \ac{GW} research in calibrating the detectors is carried out through 
a complicated system of physical manipulations of the Fabret-Perot Michelson interferometer; where 
an elaborate feedback system is used to sustain a defined measurement in arm length difference 
between the moving mirrors. A comprehensive description of the calibration procedure (in particular 
for the LIGO detectors during their fifth science run) can be found in \cite{2010NIMPA.624..223A}, 
and a brief description is giving in \cite{Vitale:2012} and references therein. For \ac{aLIGO} 
and \ac{AdV}, at the frequency range used in this project, the error in the amplitude calibration 
using the `hardware injection method' is roughly $10\%$ \cite{Vitale:2012}. This is a benchmark set 
for the error estimation using the proposed new method in this project.~\cm{Clean this up and 
expand.}

% Describe our model for the calibration
The \ac{GW} strain is measured through the differential arm length, $\Delta L$, changes of the 
interferometer via $h(f,t) = \Delta L(f,t) / L$, where $L$ is the full arm length. Calibration is 
required to relate the actual measured interferometer error signal output $e(f)$ to $\Delta L$. 
This relation is known as the length response function, $R(f)$, defined such that
\begin{equation}
\Delta L(f,t) = R(f) e(f, t),
\end{equation}
where where we assume $R$ varies only slowly in time (in comparison to transient signal timescales).
Calculation of $R$ requires the measurement of various functions within a control feedback loop
(see \cite{Vitale:2012} for a brief overview of this), which are subject to uncertainties. In this 
study we will assume an estimate of $R$ is available (although in theory we could take on the role 
of estimating $R$ itself), but that it differs from the truth through some unknown scale factor, 
$s$, so that
\begin{equation}
sh(f) = h_{\rm m}(f) = \frac{R(f,t) e(f)}{L},
\end{equation}
where $h(f)$ is the true strain and $h_{\rm m}(f)$ is the measured strain. With this definition it
means that if $s > 1$ then a signal would appear to have a larger amplitude (e.g.\ be closer) than 
in reality\footnote{However, the \ac{SNR} would be the correct value as the signal and the noise 
will be contain the scale factor}, whereas if $s < 1$ it would appear to have a lower amplitude 
than reality. In this analysis we will simplify the situation by assuming that $s$ is a constant 
with respect to frequency, but in future studies that assumption could be dropped and $s$ could 
take some functional form, or piece-wise fit, with respect to $f$.

@@

% observing a coincident event
The most likely scenario in which a coincident \ac{GW}--\ac{sGRB} event would be identified is 
through the targeted follow-up of an observed \ac{sGRB} or by post-facto matching of \ac{GW} 
trigger lists with known \ac{sGRB} events.  The likelihood of being able to follow-up \ac{GW} 
events using gamma-ray telescopes with low enough latency to catch a \ac{sGRB} is low.  Compounding 
this issue is the relatively large \ac{GW} sky error-box giving a field-of-view for the \ac{EM} 
observatories to search spanning $\sim 100$'s of square degrees~\cite{grb}.  For the \ac{GW} 
follow-up of \ac{sGRB} scenario the merger time for \ac{BNS} systems will be estimated from the 
\ac{sGRB} to within a few seconds~\cite{grb}.  This makes the follow-up search less computationally
expensive since it is performed over a smaller range of data using potentially fewer numbers of 
waveform templates.  This computational saving enables the use of a more computationally expensive 
multi-detector coherent scheme rather than the cheaper coincidence methods used in the untargeted 
searches.

@@

\cm{This should be the simplest case we can do.  We need to cut a lot of this.}

We want to assess how well the calibration scale factor can be estimated from a single observed 
\ac{GW} associated with a particular \ac{sGRB}. As we {\it a priori} have no knowledge of the 
likely location and distance of such an event we have performed simulations of multiple events to 
see the expected distribution the accuracy of the calibration scale factor recovery.

@@

For a non-spinning system the general form of the frequency-domain polarisation amplitudes are
%
\begin{eqnarray}\label{eq:signal}
  \gws_{+}(f) &\propto \frac{1+\cos^{2}\iota}{D_{L}}
\mathcal{M}^{5/6}f^{-7/6}e^{-i\Psi(f, \nu, t_\mathrm{c}, \phi_{\mathrm{c}})} \nonumber \\
  \gws_{\times}(f) &\propto
 \frac{\cos^{2}\iota}{D_{L}}\mathcal{M}^{5/6} f^{-7/6}e^{-i\Psi(f, \nu, t_\mathrm{c}, 
\phi_{\mathrm{c}})}
\end{eqnarray}
%
where the chirp-mass $\mathcal{M}$ is defined as $\mathcal{M}=M\eta^{3/5}$, with 
$\eta=m_{1}m_{2}/M^{2}$ and $M=m_{1}+m_{2}$, $\iota$ is the inclination angle, and $\nu=(\pi 
Mf)^{1/3}$.  The \ac{GW} strain measured at the $k^{\rm{th}}$ detector is then given by
%
\begin{equation}
  \label{eq:gravsig}
   \gws^{k}(f) = \big[ F_{+}^{k}(\alpha, \delta, \psi)\gws_{+}(f) +
F_{\times}^k(\alpha, \delta, \psi)\gws_{\times}(f)\big]
\end{equation}
%
where $F_{+},F_{\times}$ are the antenna response functions which are dependent upon the 
polarisation angle $\psi$ and the sky position of the source $\alpha$ and $\delta$.

% Define the detectors and the noise epoch we will be using
For this analysis we will consider the advanced generation LIGO and Virgo detectors operating
at their design sensitivity, with noise power spectral densities taken from
\cite{2013arXiv1304.0670L}. The network we consider is the two \ac{aLIGO} detectors (H1 and L1) and 
the \ac{AdV} detector (V1).

% Define the likelihood function we will be using
We consider that the measured data in any detector is defined as
%
\begin{equation}
  \tilde{d}_{k}(f)=\tilde{n}_{k}(f)+s\gws_{k}(f,\btheta)
\end{equation}
%
where $\tilde{n}(f)$ is the Fourier transform of Gaussian noise drawn from a distribution 
consistent with the \ac{PSD} of the advanced detector noise.

@@

To assess the ability to estimate the unknown calibration scale factors we want to calculate their 
marginal posterior probability distribution functions (pdf) given the data contains a signal. We 
use Bayes' theorem for which we need to define a likelihood function and prior pdfs on the signal 
parameters being estimated. In this analysis we use the Markov chain Monte Carlo code {\tt emcee} 
\cite{2013PASP..125..306F} to sample the posterior distribution. We use a Gaussian likelihood 
function given by
%
%\begin{equation}
%  p(\vec{d} | \vec{\theta}, \curlH, I) \propto \exp\Bigg[ 2\Delta
%f\sum_{k>0}\frac{|\tilde{d}(f_{k}) - \tilde{h}(f_{k})|^{2}}{S(f_{k})}\Bigg]
%\end{equation}
\begin{equation}
\fl p(\bi{d} | \btheta, \curlH, I) \propto \exp{\left[4\Delta f \sum_{k=1}^{N_{\rm det}}
  \sum_{i=i_{f_{\rm low}}}^{i_{f_{\rm high}}}\frac{\Re{\left\{ \tilde{d}^{*}_{k,i} 
\gws_{k,i}(\btheta) - \frac{1}{2}\left(\gws^*_{k,i}(\btheta)\gws_{k,i}(\btheta) + 
\tilde{d}^*_{k,i}\tilde{d}_{k,i}\right) \right\}}}{S_{k,i}}\right]}
\end{equation}
% 
where $\bi{d}$ is an array containing the data for all detectors, $\Delta f$ is the frequency 
resolution, $S_k$ is the $k^{\rm th}$ detector's one sided noise \ac{PSD}, and the $i$ indices 
increment over frequency with $i_{f_{\rm low}}$ and $i_{f_{\rm high}}$ corresponding to the lower 
and upper range in frequencies used for our analysis, which were 20\,Hz to 400\,Hz respectively. 
This frequency range was chosen as the vast majority of the \ac{SNR} for the inspiralling signals 
we use can be found within this range, so going to lower or higher frequencies provides very little 
additional information. The source parameters, and the unknown scale factors, have been subsumed 
into a vector $\btheta$.

@@

Our prior on the system component masses is dependent on whether we are considering the source 
being a \ac{BNS} system or a \ac{NSBH}. For the former case the prior we use is a Gaussian 
distribution for both components with means of $1.35\,\textrm{M}_{\odot}$ and standard deviations 
of $0.13\,\textrm{M}_{\odot}$ \MP{(reference)}. In reality, for all the systems we use in our 
study, the chirp mass, which we see from equation~\ref{eq:signal} plays a role in the overall 
signal amplitude, is very well constrained by the system's phase evolution, so our prior could be 
expanded with minimal effect on the results. In the case of \ac{NSBH} systems we use the same 
prior as above for the neutron star, but for the black hole we use a prior based on the 
canonical masss distribution used for the rate results in \cite{2012PhRvD..85h2002A} with a mean of 
$5\,\textrm{M}_{\odot}$ and a standard deviation of $1\,\textrm{M}_{\odot}$. For the \ac{NSBH}
systems we also require a prior on the black hole spin (for \ac{BNS} systems we have assumed that 
the spins are small enough that they will be negligible). We use a uniform prior on the 
normalised aligned spin magnitude between $-1$ and $1$.

@@

% Describe the ensemble of simulations used
Separately for the \ac{BNS} and \ac{NSBH} systems we have performed up to 2000 simulations at each 
distance value\footnote{Simulations on which our MCMC did not produce converged chains within our 
analysis time were discarded and not re-run, although there were always at least 1000 simulations 
that completed.}. For proposing injections for each simulation the source parameters were randomly 
drawn from the prior distributions given in Section~\ref{sec:priors}, with the sky position being 
drawn randomly from a uniform distribution, the coalescence times held fixed within the centre of 
the prior window, and with the calibration scaling factors for each detector drawn from a Gaussian 
distribution with a mean of one and standard deviation of $0.125$ (equivalent to a mean fractional 
offset of 10\%). However, in accepting a proposed injection as one to be analysed we introduced a 
criteria that the signal be ``detectable'', based on that used in \cite{2012PhRvD..85h2002A} of it 
having an \ac{SNR} of $\geq 5.5$ in at least two detectors (as these would be signals coincident 
with a \ac{GRB} we do not include the often used further constraint that the total coherent network 
\ac{SNR} is greater than 12).  This criteria means that at larger distances our population of 
injected sources was not be uniformly distributed over the sky, and were generally closer to being 
circularly polarised.

@@

For each of the simulated sources we have used the {\tt emcee} python MCMC package 
\cite{2013PASP..125..306F} to perform parameter estimation over the unknown source parameters using 
the priors as discussed in Section~\ref{sec:priors}. In each case when calculating the likelihood 
we used an estimate of the noise \ac{PSD} based on the advanced detector design sensitivities 
(using those given in \cite{2013arXiv1304.0670L}), but simulated as if calculated by averaging 64 
separate noisy \ac{PSD} estimates and scaled with the same factor as applied to the injection and 
noise\footnote{We do not account for there being a difference between the estimated \ac{PSD} and the 
actual \ac{PSD} of the analysed section of data. This difference would be very much correlated with 
the scale factor, so in reality our estimate of the scale factor would be a combination of the 
calibration offset and the difference in the \ac{PSD}. As such our results on real data would be an 
upper limit on the calibration scale factor.}. This has provided posterior probability distributions 
on the calibration scale factors for each detector. Examples of the marginalised posterior 
probabilities for a \ac{BNS} system and a \ac{NSBH} system observed with the three detector network 
are shown in figures~\ref{fig:bnspost} and \ref{fig:nsbhpost} respectively.

From these posterior distributions we have calculated the 68\% credible region on the 
calibration scale factors for each detector (if these were Gaussian distributions this is 
equivalent to the region bounded by $1\sigma$ intervals). For all the signals at each distance 
increment we have produced the distribution of the fractional half widths (i.e.\ $1\sigma$) of 
these scale factor's confidence intervals compared to the actual value. These are shown as boxplots 
in figures~\ref{fig:bnsresults} and \ref{fig:nsbhresults} for the \ac{BNS} and \ac{NSBH} systems 
respectively. The boxes show the extent from the lower to upper quartile of the values, whilst the 
whiskers represent the range of values. The black line within each box gives the median value and 
the star gives the mean value. Also shown on each plot as the dashed magenta line is the percentage 
of signal drawn from the prior distribution that would fulfill the detection criterion.

@@

\begin{figure}
 \begin{center}
  \includegraphics[width=1.0\textwidth]{scale_factor.pdf}
 \end{center}
 \caption{\label{fig:bnsresults} Distribution of the percentage accuracy at which the calibration 
scale factors can be determined for a three detector network if assuming \ac{BNS} systems (provided 
a coincident \ac{GRB} is observed and can yield a distance estimate). The boxplots span the lower 
to upper quartile range of the distributions, with the median value shown as a horizontal line 
within the box and the mean shown as a star. The dashed magenta line shows the percentage of source 
drawn from the prior distribution that would be detectable at each distance value.}
\end{figure}

\begin{figure}
 \begin{center}
  \includegraphics[width=1.0\textwidth]{scale_factor_nsbh.pdf}
 \end{center}
 \caption{\label{fig:nsbhresults} Distribution of the percentage accuracy at which the calibration 
scale factors can be determined for a three detector network if assuming \ac{NSBH} systems 
(provided a coincident \ac{GRB} is observed and can yield a distance estimate). The plot contents 
are the same as in figure~\ref{fig:bnsresults}.}
\end{figure}

In figure~\ref{fig:bnsresults} we see that on average the scale factor can be recovered to 
equivalent precision in both \ac{aLIGO} detectors, as would be expected, with uncertainties 
generally within the 10\% range for sources at 100\,Mpc. This is comparable to previous estimates 
of the calibration error in the initial LIGO detectors. An interesting feature is that for 
distances $\gtrsim 250$\,Mpc the upper extent of the uncertainty hits a maximum at $\sim 
25\mbox{--}30\%$, whilst the width of boxes narrows. This is due to our ``detectability'' criteria, 
in that for all distances we will only be seeing those sources with a high enough \ac{SNR} that we 
would consider them detectable. There will therefore be fewer sources with \ac{SNR} higher than 
this criteria at large distances, giving us a narrower range, and we automatically exclude those 
with \ac{SNR} that are too small thus truncating our uncertainty distribution at the upper end. 
However, this does show that on average for sources that are detectable out to 450\,Mpc we would be 
able to constrain the calibration scale factor uncertainty for the \ac{aLIGO} detectors to $\sim 
20\%$. As the detection criteria will generally come from the \ac{SNR} being high enough within the 
two \ac{aLIGO} instruments this means that the SNR in \ac{AdV} can be small and thus the ability to 
constrain the scale factor for it becomes poor (although it still provides information that the 
calibration is not grossly inaccurate). We also see that true uncertainties achievable for 
recovering the calibration scale factors are indeed very similar to the simple assumption that they 
would be given by $sim 1/{\rm SNR}$, i.e.\ at the upper end of the distribution the \ac{SNR} in 
the two \ac{aLIGO} detectors will be $\sim 5.5$, which would be expected to produce uncertainties 
of $\lesssim 20\%$. Our results are probably slightly poorer than the \ac{SNR} expectation due to 
their still be correlations with $\iota$ even within the constrained prior range, and slight 
correlations between the scale factors for H1 and L1 as seen in figure~\ref{fig:bnspost}.

>>>>

\begin{abstract}
  We present an investigation into the potential of assessing the validity of the 
calibration of gravitational wave detector outputs through the use of standard
  sirens. Such signals, as measured via gravitational wave
  observations provide an estimated luminosity distance which is
  subject to uncertainties in the calibration of the data.  If a host
  galaxy is identified for a given source then its redshift can be
  obtained and combined with current knowledge of the cosmological
  parameters.  This will yield the true luminosity distance and allow
  the direct comparison with the estimated value.  Discrepancies can
  then be used to correct the original calibration.~\cm{Update at end}
\end{abstract}

@@

% introduce GWs and BNS systems
It is expected that the advanced generation of interferometric \ac{GW} detectors will detect waves 
emitted from $O(10s)$ of \acp{CBC}. One such class of these cataclysmic events, the inspiral and
merger of \ac{BNS} systems will be detected out to a maximum range of $\approx 450$ Mpc. Assuming 
the current best estimates for the cosmological parameters, this is equivalent to a redshift 
$z\approx 0.1$. As noted by \cite{1986Natur.323..310S} \ac{CBC} systems can be used as cosmological 
distance markers, otherwise known as ``standard sirens'' (analogous to the \ac{EM} standard 
candles). Standard sirens allow ...

@@

% talk about the calibration idea
If a single \ac{GW} event is observed in coincidence with a \ac{sGRB} it may be possible to 
identify the host galaxy of the \ac{BNS} source.  With this information a spectroscopic redshift 
can be very accurately obtained. With knowledge of the redshift, using current best estimates for 
the Hubble constant and other cosmological parameters, the true luminosity distance can be
estimated to $\sim 1\%$ accuracy.  The \ac{GW} measurement acts as a standard siren also giving us 
a direct measurement of the luminosity distance to the source. The accuracy of such a measurement 
depends on a number of factors including the accuracy with which the \ac{GW} detector has been 
calibrated.  Hence comparison with the distance estimate from the \ac{sGRB} we can re-calibrate (or 
validate) the existing experimentally obtained calibration.

% What do we do in this paper
In this paper we investigate the feasibility of this approach for single coincident 
\ac{GW}--\ac{sGRB} events and establish the validation power of such a calibration technique.  In 
Sec.~\ref{sec:calibration} we briefly summarize the existing experimental technique for \ac{GW} 
detector calibration and its expected accuracy.  We then review the concept of \ac{GW} standard 
sirens and their proposed coincident \ac{EM} signatures, the \acp{sGRB} in
Secs.~\ref{sec:sirens} and~\ref{sec:GRB}. In Sec.~\ref{sec:cosmo} we show the likely accuracy of 
\ac{sGRB} distances obtained from their host galaxy redshift and in Sec.~\ref{sec:discussion} we 
conclude.

@@

The technique used throughout \ac{GW} research in calibrating the detectors is carried out through 
a complicated system of physical manipulations of the Fabret-Perot Michelson interferometer; where 
an elaborate feedback system is used to sustain a defined measurement in arm length difference 
between the moving mirrors. A comprehensive description of the calibration procedure (in particular 
for the LIGO detectors during their fifth science run) can be found in \cite{2010NIMPA.624..223A}, 
and a brief description is giving in \cite{Vitale:2012} and references therein. For \ac{aLIGO} 
and \ac{AdV}, at the frequency range used in this project, the error in the amplitude calibration 
using the `hardware injection method' is roughly $10\%$ \cite{Vitale:2012}. This is a benchmark set 
for the error estimation using the proposed new method in this project.

% Describe our model for the calibration
The \ac{GW} strain is measured through the differential arm length, $\Delta L$, changes of the 
interferometer via $h(f,t) = \Delta L(f,t) / L$, where $L$ is the full arm length. Calibration is 
required to relate the actual measured interferometer error signal output $e(f)$ to $\Delta L$. 
This relation is known as the length response function, $R(f)$, defined such that
\begin{equation}
\Delta L(f,t) = R(f) e(f, t),
\end{equation}
where where we assume $R$ varies only slowly in time (in comparison to transient signal 
time-scales). Calculation of $R$ requires the measurement of various functions within a control 
feedback loop (see \cite{2010NIMPA.624..223A}, which are subject to uncertainties. In this 
study we will assume an estimate of $R$ is available (although in theory we could take on the role 
of estimating $R$ itself), but that it differs from the truth through some unknown scale factor, 
$s$, so that
\begin{equation}\label{eq:scalefactor}
sh(f) = h_{\rm m}(f) = \frac{R(f,t) e(f)}{L},
\end{equation}
where $h(f)$ is the true strain and $h_{\rm m}(f)$ is the measured strain. With this definition it
means that if $s > 1$ then a signal would appear to have a larger amplitude (e.g.\ be closer) than 
in reality\footnote{However, the \ac{SNR} would be the correct value as the signal and the noise 
will both contain the scale factor}, whereas if $s < 1$ it would appear to have a lower amplitude 
than reality. In this analysis we will simplify the situation by assuming that $s$ is a constant 
with respect to frequency, but in future studies that assumption could be dropped and $s$ could 
take some functional form, or piece-wise fit, with respect to $f$ (e.g.\ in a similar way to the 
method of \cite{2013PhRvD..88h4044L} used for fitting differences in \ac{PSD} estimates).

@@

% observing a coincident event
The most likely scenario in which a coincident \ac{GW}--\ac{sGRB} event would be identified is 
through the targeted follow-up of an observed \ac{GW} or by post-facto matching of \ac{GW} 
trigger lists with known \ac{sGRB} events.  The likelihood of being able to follow-up \ac{GW} 
events using gamma-ray telescopes with low enough latency to catch a \ac{sGRB} is low.  Compounding 
this issue is the relatively large \ac{GW} sky error-box giving a field-of-view for the \ac{EM} 
observatories to search spanning $\sim 100$'s of square degrees~\cite{grb}.  For the \ac{GW} 
follow-up of \ac{sGRB} scenario the merger time for \ac{BNS} systems will be estimated from the 
\ac{sGRB} to within a few seconds~\cite{grb}.  This makes the follow-up search less computationally
expensive since it is performed over a smaller range of data using potentially fewer numbers of 
waveform templates.  This computational saving enables the use of a more computationally expensive 
multi-detector coherent scheme rather than the cheaper coincidence methods used in the untargeted 
searches.

@@

We want to assess how well the calibration scale factor defined in Eqn.~\ref{eq:scalefactor} can be 
estimated from a single observed \ac{GW} associated with a particular \ac{sGRB}. As we {\it a 
priori} have no knowledge of the likely location and distance of such an event we have performed 
simulations of multiple events to see the expected distribution in the accuracy of the calibration 
scale factor recovery.

@@

For a non-spinning system the general form of the frequency-domain polarisation amplitudes are
%
\begin{eqnarray}\label{eq:signal}
  \gws_{+}(f) &\propto \frac{1+\cos^{2}\iota}{D_{L}}
\mathcal{M}^{5/6}f^{-7/6}e^{-i\Psi(f, \nu, t_\mathrm{c}, \phi_{\mathrm{c}})} \nonumber \\
  \gws_{\times}(f) &\propto
 \frac{\cos^{2}\iota}{D_{L}}\mathcal{M}^{5/6} f^{-7/6}e^{-i\Psi(f, \nu, t_\mathrm{c}, 
\phi_{\mathrm{c}})}
\end{eqnarray}
%
where the chirp-mass $\mathcal{M}$ is defined as $\mathcal{M}=M\eta^{3/5}$, with 
$\eta=m_{1}m_{2}/M^{2}$ and $M=m_{1}+m_{2}$, $\iota$ is the inclination angle, and $\nu=(\pi 
Mf)^{1/3}$.  The \ac{GW} strain measured at the $k^{\rm{th}}$ detector is then given by
%
\begin{equation}
  \label{eq:gravsig}
   \gws^{k}(f) = \big[ F_{+}^{k}(\alpha, \delta, \psi)\gws_{+}(f) +
F_{\times}^k(\alpha, \delta, \psi)\gws_{\times}(f)\big]
\end{equation}
%
where $F_{+}$ and $F_{\times}$ are the antenna response functions which are dependent upon the 
polarisation angle $\psi$ and the sky position of the source $\alpha$ and $\delta$.

% Define the detectors and the noise epoch we will be using
For this analysis we will consider the advanced generation \ac{aLIGO} and \ac{AdV} detectors 
operating at their design sensitivity, with noise power spectral densities taken from
\cite{2013arXiv1304.0670L}. The network we consider is the two \ac{aLIGO} detectors (H1 and L1) and 
the \ac{AdV} detector (V1).

% Define the likelihood function we will be using
We consider that the measured data in any detector is defined as
%
\begin{equation}
  \tilde{d}_{k}(f)=s(\tilde{n}_{k}(f)+\gws_{k}(f,\btheta))
\end{equation}
%
where $\tilde{n}(f)$ is the Fourier transform of the true (Gaussian) strain noise drawn from a 
distribution consistent with the \ac{PSD} of the advanced detector noise and $\btheta$ is the set 
of waveform parameters.

@@

To assess the ability to estimate the unknown calibration scale factors we want to calculate their 
marginal posterior probability distribution functions (pdf). We use Bayes' theorem for which we 
need to define a likelihood function and prior pdfs on the signal parameters being estimated. In 
this analysis we use the Markov chain Monte Carlo code {\tt emcee} \cite{2013PASP..125..306F} to 
sample the posterior distribution. We use a Gaussian likelihood function given by
%
\begin{equation}
\fl p(\bi{d} | \btheta, \bi{s}, \curlH, I) \propto \exp{\left[4\Delta f \sum_{k=1}^{N_{\rm det}}
  \sum_{i=i_{f_{\rm low}}}^{i_{f_{\rm high}}}\frac{\Re{\left\{ \tilde{d}^{*}_{k,i} 
\gws_{k,i}(\btheta) - \frac{1}{2}\left(\gws^*_{k,i}(\btheta)\gws_{k,i}(\btheta) + 
\tilde{d}^*_{k,i}\tilde{d}_{k,i}\right) \right\}}}{S_{k,i}}\right]}
\end{equation}
% 
where $\bi{d}$ is an array containing the data for all detectors, $\Delta f$ is the frequency 
resolution, $S_k$ is the $k^{\rm th}$ detector's one sided noise \ac{PSD}, and the $i$ indices 
increment over frequency with $i_{f_{\rm low}}$ and $i_{f_{\rm high}}$ corresponding to the lower 
and upper range in frequencies used for our analysis, which were 20\,Hz to 400\,Hz respectively. 
This frequency range was chosen as the vast majority of the \ac{SNR} for the inspiralling signals 
we use can be found within this range, so going to lower or higher frequencies provides very little 
additional information. The source parameters, and the unknown scale factors, have been subsumed 
into a vector $\btheta$.

@@

Our prior on the system component masses is dependent on whether we are considering the source 
being a \ac{BNS} system or a \ac{NSBH}. For the former case the prior we use is a Gaussian 
distribution for both components with means of $1.35\,\textrm{M}_{\odot}$ and standard deviations 
of $0.13\,\textrm{M}_{\odot}$ \MP{(reference)}. In reality, for all the systems we use in our 
study, the chirp mass, which we see from equation~\ref{eq:signal} plays a role in the overall 
signal amplitude, is very well constrained by the system's phase evolution, so our prior could be 
expanded with minimal effect on the results. In the case of \ac{NSBH} systems we use the same 
prior as above for the neutron star, but for the black hole we use a prior based on the 
canonical masss distribution used for the rate results in \cite{2012PhRvD..85h2002A} with a mean of 
$5\,\textrm{M}_{\odot}$ and a standard deviation of $1\,\textrm{M}_{\odot}$. We note that the mass 
range of black holes could be quite different from this, but use this range as an example for this 
type of system. For the \ac{NSBH} systems we also require a prior on the black hole spin (for 
\ac{BNS} systems we have assumed that the spins are small enough that they will be negligible). We 
use a uniform prior on the normalised aligned spin magnitude between $-1$ and $1$.

@@

% Describe the ensemble of simulations used
Separately for the \ac{BNS} and \ac{NSBH} systems we have performed a large number simulations at 
each distance value. For proposing injections for each simulation the source parameters were 
randomly drawn from the prior distributions given in Section~\ref{sec:priors}, with the sky 
position being drawn randomly from a uniform distribution, the coalescence times held fixed within 
the centre of the prior window, and with the calibration scaling factors for each detector drawn 
from a Gaussian distribution with a mean of one and standard deviation of $0.125$ (equivalent to a 
mean fractional offset of 10\%). However, in accepting a proposed injection as one to be analysed 
we introduced a criteria that the signal be ``detectable'', based on that used in 
\cite{2012PhRvD..85h2002A} of it having an \ac{SNR} of $\geq 5.5$ in at least two detectors (as 
these would be signals coincident with a \ac{sGRB} we do not include the often used further 
constraint that the total coherent network \ac{SNR} is greater than 12). This criteria means that 
at larger distances our population of injected sources was not uniformly distributed over the sky, 
and were generally closer to being circularly polarised.

@@

For each of the simulated sources we have used the {\tt emcee} python MCMC package 
\cite{2013PASP..125..306F} to perform parameter estimation over the unknown source parameters using 
the priors as discussed in Section~\ref{sec:priors}. In each case when calculating the likelihood 
we used an estimate of the noise \ac{PSD} based on the advanced detector design sensitivities 
(using those given in \cite{2013arXiv1304.0670L}), but simulated as if calculated by averaging 64 
separate noisy \ac{PSD} estimates and scaled with the same factor as applied to the injection and 
noise\footnote{We do not account for there being a potential difference between the estimated 
\ac{PSD} and the actual \ac{PSD} of the analysed section of data as described in e.g.\ 
\cite{2013PhRvD..88h4044L}. This difference would be very highly correlated with the calibration 
scale factor, so in reality our estimate of the scale factor would be a combination of the
calibration offset and the difference in the \ac{PSD}. As such our results on real data would be an 
upper limit on the calibration scale factor.}. This has provided posterior probability distributions 
on the calibration scale factors for each detector. Examples of the marginalised posterior 
probabilities for a \ac{BNS} system and a \ac{NSBH} system observed with the three detector network 
are shown in figures~\ref{fig:bnspost} and \ref{fig:nsbhpost} respectively.

From these posterior distributions we have calculated the 68\% credible region on the calibration 
scale factors for each detector (if these were Gaussian distributions this is equivalent to the 
region either side of the mean bounded by the $1\sigma$ intervals). For all the signals at each 
distance increment we have produced the distribution of the fractional half widths (i.e.\ $1\sigma$) 
of these scale factors' confidence intervals compared to the actual value. These are shown as 
boxplots in figures~\ref{fig:bnsresults} and \ref{fig:nsbhresults} for the \ac{BNS} and \ac{NSBH} 
systems respectively. The boxes show the extent from the lower to upper quartile of the values, 
whilst the whiskers extend from the 5$^{\rm th}$ to 95$^{\rm th}$ percentile. The black line 
within each box gives the median value and the star gives the mean value. Also shown on each plot as 
the dashed magenta line is the percentage of signals drawn from the prior distribution that fulfill 
the detection criterion.

@@

\begin{figure}
 \begin{center}
  \includegraphics[width=1.0\textwidth]{scale_factor.pdf}
 \end{center}
 \caption{\label{fig:bnsresults} Distributions of the percentage accuracy at which the calibration 
scale factors can be determined for a three detector network if assuming \ac{BNS} systems (provided 
a coincident \ac{GRB} is observed and can yield a distance estimate). The boxplots span the lower 
to upper quartile range of the distributions, with the median value shown as a horizontal line 
within the box and the mean shown as a star. The dashed magenta line shows the percentage of source 
drawn from the prior distribution that would be detectable at each distance value.}
\end{figure}

\begin{figure}
 \begin{center}
  \includegraphics[width=1.0\textwidth]{scale_factor_nsbh.pdf}
 \end{center}
 \caption{\label{fig:nsbhresults} Distributions of the percentage accuracy at which the calibration 
scale factors can be determined for a three detector network if assuming \ac{NSBH} systems 
(provided a coincident \ac{GRB} is observed and can yield a distance estimate). The plot contents 
are the same as in figure~\ref{fig:bnsresults}.}
\end{figure}

In figure~\ref{fig:bnsresults} we see that on average the scale factor can be recovered to 
equivalent precision in both \ac{aLIGO} detectors, as would be expected, with uncertainties 
generally within the 10\% range for sources at 100\,Mpc. This is comparable to previous estimates 
of the calibration error in the initial LIGO detectors. An interesting feature is that for 
distances $\gtrsim 250$\,Mpc the upper extent of the uncertainty for H1 and L1 hits a maximum at 
$\sim 25\%$, whilst the width of boxes narrows. This is due to our ``detectability'' criteria, 
in that for all distances we will only be seeing those sources with a high enough \ac{SNR} that we 
would consider them detectable. There will therefore be fewer sources with \ac{SNR} higher than 
this criteria at large distances, giving us a narrower range, and we automatically exclude those 
with \ac{SNR} that are too small thus truncating our uncertainty distribution at the upper end. 
However, this does show that on average for sources that are detectable out to 450\,Mpc we would be 
able to constrain the calibration scale factor uncertainty for the \ac{aLIGO} detectors to $\sim 
20\%$. As the detection criteria will generally come from the \ac{SNR} being high enough within the 
two \ac{aLIGO} instruments this means that the SNR in \ac{AdV} can be small and thus the ability to 
constrain the scale factor for it becomes poor (although it still provides information that the 
calibration is not grossly inaccurate). We also see that true uncertainties achievable for 
recovering the calibration scale factors are indeed very similar to the simple assumption that they 
would be given by $sim 1/{\rm SNR}$, i.e.\ at the upper end of the distribution the \ac{SNR} in 
the two \ac{aLIGO} detectors will be $\sim 5.5$, which would be expected to produce uncertainties 
of $\lesssim 20\%$. Our results are probably slightly poorer than the \ac{SNR} expectation due to 
their still be correlations with $\iota$ even within the constrained prior range, and slight 
correlations between the scale factors for H1 and L1 as seen in figure~\ref{fig:bnspost}.
